---
title: "Week 5: Data Wrangling"
subtitle: "PM 566: Introduction to Health Data Science"
author: "George G. Vega Yon"
output:
  slidy_presentation:
    slide_level: 1
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(eval = TRUE, results = "hold")
slides_eval <- FALSE
```


<!--Yeah... I have really long code chunks, so I just changed the default size :)-->
<style type="text/css">
code.*, .remark-code, pre {
  font-size:16px;
}
p,h1,h2,h3,h4 {
  font-family: system-ui
}
</style>


## Today's goals

We will learn about how to manipulate data, and in particular,

- Selecting variables.
- Filtering data.
- Creating variables.
- Summarize data.

Throughout the session we will see examples using: 

- [**data.table**](https://cran.r-project.org/package=data.table) in R,
- [**dtplyr**](https://cran.r-project.org/package=dplyr) in R, and
- [**pydatatable**](https://github.com/h2oai/datatable)

All with the [MET](https://github.com/USCbiostats/data-science-data/) dataset.

---

## Data wrangling in R

Overall, you will find the following approaches:

- **base R**: Use only base R functions.

- **dplyr**: Using "verbs".

- **data.table**: High-performing (ideal for large data)

- **dplyr + data.table = dtplyr**: High-performing + dplyr verbs.

Other methods involve, for example, using external tools such as [Spark](http://spark.apache.org/), [sparkly](https://cran.r-project.org/package=sparklyr).

We will be focusing on data.table because of [this](https://h2oai.github.io/db-benchmark/)

Take a look at this very neat cheat sheet by
[Erik Petrovski](http://www.petrovski.dk/) [here](https://github.com/rstudio/cheatsheets/raw/master/datatable.pdf).

---

## Selecting variables: Load the packages

```{r loading-packages, eval = TRUE}
library(data.table)
library(dtplyr)
library(dplyr)
```

The `dtplyr` R package translates `dplyr` (`tidyverse`) syntax to `data.table`,
so that we can still use **the dplyr verbs** while at the same time leveraging the
performance of `data.table`.

---

## Loading the data

The data that we will be using is an already processed version of the MET
dataset. We can download (and load) the data directly in our session using the
following commands:

```{r downloading-data, cache=slides_eval}
# Where are we getting the data from
met_url <- "https://github.com/USCbiostats/data-science-data/raw/master/02_met/met_all.gz"

# Downloading the data to a tempfile (so it is destroyed afterwards)
# you can replace this with, for example, your own data:
# tmp <- tempfile(fileext = ".gz")
tmp <- "met.gz"

# We sould be downloading this, ONLY IF this was not downloaded already.
# otherwise is just a waste of time.
if (!file.exists(tmp)) {
  download.file(
    url      = met_url,
    destfile = tmp,
    # method   = "libcurl", timeout = 1000 (you may need this option)
  )
}
```

Now we can load the data using the `fread()` function.

---

In R

```{r reading-data, eval = TRUE, echo = TRUE, results='hide'}
# Reading the data
dat <- fread(tmp)
head(dat)
```

In Python

```{python reading-py, eval = TRUE, echo = TRUE, results='hide'}
import datatable as dt
dat = dt.fread("met.gz")
dat.head(5)
```

Before we continue, let's learn a bit more on `data.table` and `dtplyr`

---

## `data.table` and `dtplyr`:Data Table's Syntax

- As you have seen in previous lectures, in `data.table` all happens within the
square brackets. Here is common way to imagine DT:

<div align="center">
<img src="datatable-syntax-paths.svg" width="700px">
</div>

- Any time that you see **:=** in **j** that is "Assignment by reference." Using
  **=** within **j** only works in some specific cases.
  
---

## `data.table` and `dtplyr`:Data Table's Syntax

- Operations applied in **j** are evaluated *within* the data, meaning that
  names work as symbols. e.g.
  
  ```{r error-j, eval=FALSE}
  data(USArrests)
  USArrests_dt <- data.table(USArrests)
  
  # This returns an error
  USArrests[, Murder]
  
  # This works fine
  USArrests_dt[, Murder]
  ```
  
  Furthermore, we can do things like this:
  
```{r load-usarrests, echo = FALSE}
data("USArrests")
USArrests_dt <- data.table(USArrests)
```
  
  
  ```{r plot-usarrests}
  USArrests_dt[, plot(Murder, UrbanPop)]
  ```
  
  

---

## `data.table` and `dtplyr`: Lazy table

- The `dtplyr` package provides a way to translate `dplyr` verbs to `data.table` syntax.

- The key lies on the function `lazy_dt` from `dtplyr` (see `?dtplyr::lazy_dt`).

- This function creates a wrapper that "points" to a `data.table` object

---

## `data.table` and `dtplyr`: Lazy table (cont.)

```{r lazy-dt}
# Creating a lazy table object
dat_ldt <- lazy_dt(dat, immutable = FALSE)

# We can use the address() function from data.table
address(dat)
address(dat_ldt$parent)
```

<div align="center">
<img src="pointer-paths.svg" width="800px">
</div>

Question: What is the `immutable = FALSE` option used for?

---

## Selecting columns

How can we select the columns `USAFID`, `lat`, and `lon`

```{r select1-dt, eval = slides_eval}
# data.table
dat[, list(USAFID, lat, lon)]

# dplyr verb
dat_ldt %>% select(USAFID, lat, lon)
```

In the case of `data.table`, What happens if instead of `list()` you used `c()`? Now, in Python

```{python select1-py, eval = slides_eval}
dat[:,["USAFID", "lat", "lon"]]
```

What happens if instead of `["USAFID", "lat", "lon"]` you used `{"USAFID", "lat", "lon"}` (vector vs set).

---

## Data filtering: Logical conditions

- Based on logical operations, e.g. `condition 1 [and|or condition2 [and|or ...]]`

- Need to be aware of ordering and grouping of `and` and `or` operators.

- Fundamental **logical** operators:
  
  ```{r operators, echo = FALSE}
  test <- expand.grid(c(TRUE, FALSE),c(TRUE, FALSE))
  ans <- data.frame(
    x = ifelse(test[,1], "true", "false"),
    y = ifelse(test[,2], "true", "false"),
    `Negate<br>!x`       = ifelse(!test[,1] , "true", "false"),
    `And<br>x & y`       = ifelse(test[,1] & test[,2], "true", "false"),
    `Or<br>x | y`        = ifelse(test[,1] | test[,2], "true", "false"),
    `Xor<br>xor(x, y)`       = ifelse(xor(test[,1], test[,2]), "true", "false"),
    check.names     = FALSE
  )
  
  knitr::kable(ans, row.names = FALSE, align = "c")
  ```

- Fundamental **relational** operators, in R: `<`, `>`, `<=`, `>=`, `==`, `!=`.


---

## Questions 1: How many ways can you write an XOR operator?

Write a function that takes two arguments `(x,y)` and applies the XOR operator
element wise. Here you have a template:

```r
myxor <- function(x, y) {
  res <- logical(length(x))
  for (i in 1:length(x)) {
    res[i] <- # do something with x[i] and y[i]
  }
  return(res)
}
```

Or if vectorized (which would be better)

```r
myxor <- function(x, y) {
  # INSERT YOUR CODE HERE
}
```



Hint 1: Remember that negating `(x & y)` equals `(!x | !y)`.

Hint 2: Logical operators are a distributive, meaning
`a * (b + c) = (a * b) + (a + c)`, where `*` and `+` are `&` or `|`.

---

In R

```{r xor-r}
myxor1 <- function(x,y) {(x & !y) | (!x & y)}
myxor2 <- function(x,y) {!((!x | y) & (x | !y))}
myxor3 <- function(x,y) {(x | y) & (!x | !y)}
myxor4 <- function(x,y) {!((!x & !y) | (x & y))}
cbind(
  ifelse(xor(test[,1], test[,2]), "true", "false"),
  ifelse(myxor1(test[,1], test[,2]), "true", "false"),
  ifelse(myxor2(test[,1], test[,2]), "true", "false"),
  ifelse(myxor3(test[,1], test[,2]), "true", "false"),
  ifelse(myxor4(test[,1], test[,2]), "true", "false")
)
```

---

Or in python

```{python xor-py}
# Loading the libraries
import numpy as np
import pandas as pa

# Defining the data
x = [True, True, False, False]
y = [False, True, True, False]
ans = {
    'x'   : x,
    'y'   : y,
    'and' : np.logical_and(x, y),
    'or'  : np.logical_or(x, y),
    'xor' : np.logical_xor(x, y)
    }
pa.DataFrame(ans)
```

---

Or in python (bis)

```{python xor-py-bis}
def myxor(x,y):
    return np.logical_or(
        np.logical_and(x, np.logical_not(y)),
        np.logical_and(np.logical_not(x), y)
    )

ans['myxor'] = myxor(x,y)
pa.DataFrame(ans)
```

We will now see applications using the `met` dataset

---

## Filtering (subsetting) the data

Need to select records according to some criteria. For example:

- First day of the month, and
- Above latitude 40, and
- Elevation outside the range 500 and 1,000.

---

In R with `data.table`

- `(day == 1)` First day of the month, and
- `(lat > 40)` Above latitude 40, and
- `((elev < 500) | (elev > 1000))` Elevation outside the range 500 and 1,000.

```{r filter1-dt1, eval = slides_eval}
dat[(day == 1) & (lat > 40) & ((elev < 500) | (elev > 1000))] %>%
  nrow()
```

In R with `dtplyr`

```{r filter1-tv, eval = slides_eval}
dat_ldt %>%
  filter(day == 1, lat > 40, (elev < 500) | (elev > 1000)) %>%
  compute() %>% # Notice this line!
  nrow()
```

---

In Python

```{python filter1-py-load, eval = slides_eval}
import datatable as dt
dat = dt.fread("met.gz")
```

- `(dt.f.day == 1)` First day of the month, and
- `(dt.f.lat > 40)` Above latitude 40, and
- `((dt.f.elev < 500) | (dt.f.elev > 1000))` Elevation outside the range 500 and 1,000.

```{python filter1-py, eval = slides_eval}
dat[(dt.f.day == 1) & (dt.f.lat > 40) & ((dt.f.elev < 500) | (dt.f.elev > 1000)), :].nrows
# dat[dt.f.day == 1,:][dt.f.lat > 40,:][(dt.f.elev < 500) | (dt.f.elev > 1000),:].nrows
```

In the case of pydatatable we use `dt.f.` to refer to a column. `df.` is what
we use to refer to datatable's [namespace](https://en.wikipedia.org/wiki/Namespace).

The [`f.` is a symbol](https://datatable.readthedocs.io/en/latest/manual/f-expressions.html)
that allows accessing column names in a datatable's `Frame`.

---

## Questions 2

1. How many records have a temperature within 18 and 25?

2. Some records have missings. Count how many records have `temp` as `NA`?

3. Following the previous question, plot a sample of 1,000 of `(lat, lon)` of the
   stations with temp as NA and those with data.

--- 

## Solutions

```{r echo=FALSE, eval = TRUE, cache = TRUE}
# Question 1
message("Question 1: ", nrow(dat[(temp < 25) & (temp > 18)]))
# dat[temp %between% c(18, 25), .N] 

# dat_ldt %>% filter(between(temp, 18, 25)) %>% compute() %>% nrow()

# Question 2
message("Question 2: ", dat[is.na(temp), .N])

# Question 3
set.seed(123)
message("Question 3")
dat[is.na(temp)][sample.int(.N, 1000), plot(lon, lat, col = "steelblue", pch=8)]
dat[!is.na(temp)][sample.int(.N, 1000), points(lon, lat, col = "tomato", pch=8)]
legend(
  "bottomleft",
  fill = c("steelblue", "tomato"),
  legend = c("temp is NA", "temp is not NA"),
  bt = "n"
  )
```


---

## Creating variables: Data types

- **logical**: Bool true/false type, e.g. dead/alive, sick/healthy, good/bad, yes/no, etc.

- **strings**: string of characters (letters/symbols), e.g. names, text, etc.

- **integer**: Numeric variable with no decimal (discrete), e.g. age, days, counts, etc.

- **double**: Numeric variable with decimals (continuous), e.g. distance, expression level, time.

In C (and other languages), strings, integers, and doubles may be specified 
with size, e.g. in `python` integers can be of 9, 16, and 32 bits. This is relevant
when managing large datasets, where saving space can be fundamental
([more info](https://en.wikipedia.org/wiki/C_data_types#Main_types)).

---

## Creating variables: Special data types

Most programming languages have special types which are built using basic types.
A few examples:

- **time**: Could be date, date + time, or a combination of both. Usually it has
  a reference number defined as date 0. In R, the `Date` class has as reference
  1970-01-01, in other words, "days since January 1st, 1970".
  
- **categorical**: Commonly used to represent strata/levels of variables, e.g. a variable
  "country" could be represented as a factor, where the data is stored as numbers
  but has a label.
  
- **ordinal**: Similar to factor, but it has ordering, e.g.
  "satisfaction level: 5 very satisfied, ..., 1 very unsatisfied".

Other special data types could be ways to represent missings (usually described as `na` or `NA`),
or special numeric types, e.g. `+-Inf` and Undefined (`NaN`).


When storing/sharing datasets, it is a good practice to do it along a dictionary
describing each column data type/format.

---

## Questions 3: What's the best way to represent the following

- 0, 1, 1, 0, 0, 1

- Diabetes type 1, Diabetes type 2, Diabetes type 1, Diabetes type 2

- on, off, off, on, on, on

- 5, 10, 1, 15, 0, 0, 1

- 1.0, 2.0, 10.0, 6.0

- high, low, medium, medium, high

- -1, 1, -1, -1, 1,

- .2, 1.5, .8, $\pi$

- $\pi$, $\exp{1}$, $\pi$, $\pi$

---

## Variable creation

If we wanted to create two variables, `elev^2` and the scaled version of `wind.sp` by it's
standard error, we could do the following

With `data.table`

```{r eval=slides_eval}
dat[, elev2         := elev^2]
dat[, windsp_scaled := wind.sp/sd(wind.sp, na.rm = TRUE)]
# Alternatively:
# dat[, c("elev2", "windsp_scaled") := .(elev^2, wind.sp/sd(wind.sp,na.rm=TRUE)) ]
```

With dtplyr

```{r eval=slides_eval}
dat[, c("elev2", "windsp_scaled") := NULL] # This to delete these variables
dat_ldt %>%
  mutate(
    elev2         = elev ^ 2,
    windsp_scaled = wind.sp/sd(wind.sp,na.rm=TRUE)
  ) %>% compute()
```

Notice that, since `dat_ldt` is a pointer, the variables are now in `dat`.

---

## Merging data

- While building the MET dataset, we dropped the State data.

- We can use the original Stations dataset and *merge* it to the MET dataset.

- But we cannot do it right away. We need to process the data somewhat first.


---

## Merging data (cont. 1)

```{r stations-data, cache = TRUE, warning=FALSE}
stations <- fread("ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
stations[, USAF := as.integer(USAF)]

# Dealing with NAs and 999999
stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

# Selecting the three relevant columns, and keeping unique records
stations <- unique(stations[, list(USAF, CTRY, STATE)])

# Dropping NAs
stations <- stations[!is.na(USAF)]

head(stations)
```

Notice the function `fifelse()`. Now, let's try to merge the data!

---

## Merging data (cont. 2)

```{r merge-try1, cache = TRUE}
merge(
  # Data
  x     = dat,      
  y     = stations, 
  # List of variables to match
  by.x  = "USAFID",
  by.y  = "USAF", 
  # Which obs to keep?
  all.x = TRUE,      
  all.y = FALSE
  ) %>% nrow()
```

This is more rows! The original dataset, `dat`, has `r nrow(dat)`. This means
that the `stations` dataset has duplicated IDs. We can fix this:

```{r drop-dpl, results='hide'}
stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]
```

---

## Merging data (cont. 3)

We now can use the function `merge()` to add the extra data

```{r merge, eval = TRUE}
dat <- merge(
  # Data
  x     = dat,      
  y     = stations, 
  # List of variables to match
  by.x  = "USAFID",
  by.y  = "USAF", 
  # Which obs to keep?
  all.x = TRUE,      
  all.y = FALSE
  )

head(dat[, list(USAFID, WBAN, STATE)])
```

What happens when you change the options `all.x` and `all.y`?

```{r merging-all-x-y, include = FALSE}
merge(x = dat, y = stations, by.x = "USAFID", by.y = "USAF",all.x = TRUE, all.y = FALSE) %>% nrow()
merge(x = dat, y = stations, by.x = "USAFID", by.y = "USAF",all.x = TRUE, all.y = TRUE) %>% nrow()
merge(x = dat, y = stations, by.x = "USAFID", by.y = "USAF",all.x = FALSE, all.y = TRUE) %>% nrow()
merge(x = dat, y = stations, by.x = "USAFID", by.y = "USAF",all.x = FALSE, all.y = FALSE) %>% nrow()
```

---

## Aggregating data: Adding groupped variables

- Many times we need to either impute some data, or generate variables by strata.

- If we, for example, wanted to impute missing temperature with the daily state
  average, we could use **by** together with the **data.table::fcoalesce()** function:
  
  ```{r group-dt1, eval = TRUE}
  # Generating a date component
  dat[, temp_imp := fcoalesce(temp, mean(temp, na.rm = TRUE)), by = .(month, day, year, STATE)]
  ```

- In the case of dplyr, we can do the following using **dplyr::group_by** together
  with **dplyr::coalesce()**:
  
  ```{r group-tv1, eval = TRUE, results='hide'}
  # We need to create the lazy table again, since we replaced it in the merge
  dat_ldt <- lazy_dt(dat, immutable = FALSE)
  
  dat_ldt %>%
    group_by(month, day, year, STATE) %>%
    mutate(
      temp_imp2 = coalesce(temp, mean(temp, na.rm = TRUE))
      ) %>% compute()
  ```

---

## Aggregating data: Adding groupped variables (cont.)

Let's see how it looks like

```{r plot-groupped}
dat[USAFID == 720172][order(USAFID, year, month, day)][{
  op <- par(mfrow = c(1, 2))
  plot(temp, type = "l", col = "tomato")
  plot(temp_imp, type = "l", col = "steelblue")
  par(op)
}]
```

---

```{r eval=FALSE}
dat[,
  list(
    wind.sp   = mean(wind.sp, na.rm = TRUE),
    temp      = mean(temp, na.rm = TRUE),
    atm.press = mean(atm.press, na.rm = TRUE)
  ), by = c("USAFID", "Date")
  ]

head(dat)
```

---

The previous is equivalent to write (compare both using the `identical()` function in R)

```r
dat_daily2 <- dat[
  ,
  .(
    wind.sp   = mean(wind.sp, na.rm = TRUE),
    temp      = mean(temp, na.rm = TRUE),
    atm.press = mean(atm.press, na.rm = TRUE)
  ),
  by = .(USAFID, Date)
]
```

---

When dealing with too many variables, we can use the `.SD` special symbol:


```{r group-dt2, eval = FALSE}
# Listing the names
in_names  <- c("wind.sp", "temp", "atm.press")
out_names <- paste0(in_names, "_avg")

dat[,
  c(out_names) := lapply(.SD, mean, na.rm = TRUE),
  .SDcols = in_names,
  by      = c("USAFID", "Date")
  ]
```

---

## Other data.table goodies

- `shift()` Fast lead/lag for vectors and lists.

- `fifelse()` Fast if-else, similar to base R's `ifelse()`.

- `fcoalesce()` Fast coalescing of missing values.

- `%between%` A short form of `(x < lb) & (x > up)`

- `%inrange%` A short form of `x %in% lb:up`

- `%chin%` Fast match of character vectors, equivalent to `x %in% X`,
  where both `x` and `X` are character vectors.
  
- `nafill()` Fill missing values using a constant, last observed value, or
  the next observed value.

---

## Scenarios

1. The 9, 99, 999, ... coding for missigness.

2. The data contains dates

3. The data is a panel dataset with various measures per unit, and we need to create a lag

4. Need to create some summaries at the group level

5. Have two datasets and need to merge them by one or more keys/indices/ids.

6. Create a running mean spanning the last 7 observations.

---

## Data Management: data.table

The data.table package:

- Is very fast
- Is memory efficient
- Is 

---

## Benchmarks

- [H2O.ai](https://www.h2o.ai/)'s benchmark ([link](https://h2oai.github.io/db-benchmark/)): Designed by
  the lead developer of data.table [Matt Dowle](https://rdatatable.gitlab.io/data.table/)

- [RStudio's](https://rstudio.com) benchmark ([link](https://cran.r-project.org/web/packages/vroom/vignettes/benchmarks.html)):
  Designed as part of the benchmarks with the [vroom](https://cran.r-package.org/package=vroom) package.

---

```{r eval=FALSE}
# Monthly average temperature plot
library(ggplot2)
library(akima)
dat_daily_plot <- with(
  dat_daily[mday(Date) == 1 & is.finite(temp_avg)],
  akima::interp(
    x = lon, y = lat, z = temp_avg
  )
)

image(dat_daily_plot)
```

---

```{r, eval = FALSE}
library(dtplyr)
library(dplyr)
data(diamonds, package="ggplot2")
dt_tibble <- lazy_dt(diamonds)
dt_dt     <- lazy_dt(as.data.table(diamonds))
dt        <- as.data.table(diamonds)
microbenchmark::microbenchmark(
  dplyr         = diamonds %>% group_by(carat) %>% summarise(mean(price), .groups="drop") %>% collect(),
  dtplyr_tibble = dt_tibble %>% group_by(carat) %>% summarise(mean(price)) %>% collect(),
  dtplyr_dt     = dt_dt %>% group_by(carat) %>% summarise(mean(price), .groups="drop") %>% collect(),
  data.table    = dt[, mean(price), by = carat]
)

```

```{r, eval = FALSE}
tmp <- tempfile(fileext="csv")
download.file(
  "https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv",
  destfile = tmp,
  method   = "libcurl",
  timeout  = 1000
  )
dat_dt <- fread(tmp)
dat_tb <- as_tibble(dat_dt)
dat_lz <- lazy_dt(dat_tb)
```


```{r, eval = FALSE}
microbenchmark::microbenchmark(
  dplyr         = dat_tb %>% group_by(dest) %>% summarise(mean(air_time), .groups="drop") %>% collect(),
  dtplyr_tibble = dat_lz %>% group_by(dest) %>% summarise(mean(air_time), .groups="drop") %>% collect(),
  data.table    = dat_dt[, mean(air_time), by = dest], times = 100
)
```


