---
output: github_document
---

# Course Material for PM566

The PM566 course website is [here](https://elastic-khorana-70231e.netlify.app/). We will be using this website to share all course material including lecture slides, labs, and assignments.

Our class meets at 1pm every Wednesday. Zoom links for each lecture session are available in the course Blackboard.

# Latest week's cross reference

This is an example of web scrapping. We built this program to extract the
cross-reference links from a given GitHub issue, including details such as timestamp,
the user, and the commit message.

You can download the program [here](list_cross_ref.R)

```{r}
library(data.table)
library(stringr)

# API access
source("list_cross_ref_api.R", echo=FALSE)
```


```{r webscraping, message=FALSE, cache=TRUE}
library(httr)
# Setting up the token
if (file.exists(".github_app.R")) {
  # See
  # https://github.com/r-lib/httr/blob/cb4e20c9e0b38c0c020a8756db8db7a882288eaf/demo/oauth2-github.r
  # For a detailed example
  source(".github_app.R")
  token <- oauth2.0_token(
    endpoint = oauth_endpoints("github"),
    app      = github_app,
    config   = httr::config(connecttimeout = 60)
    )
  token <- httr::config(token = token)
} else {
  token <- NULL
}

# Retrieving all the cross-reference information
ans <- list_cross_ref_api(
  issue = 26,
  config = c(httr::config(connecttimeout=60), token)
  )
```


```{r printing-the-data, message=FALSE}
# Preparing to print using a nice format
ans[, user := sprintf("[%s](https://github.com/%s)", gsub("/.+", "", repo), repo)]
ans[, link := sprintf("[link](%s)", commit_url)]
setorder(ans, created_at)

# A more concise message
ans[, message := sprintf("%s (%s to this %s) (files: %s)", message, link, type, files)]
knitr::kable(ans[, .(created_at, user, message, type)])
```

