---
title: "Untitled"
output:
  xaringan::moon_reader:
    css: ["theme.css", "default"]
    lib_dir: libs
    nature:
      beforeInit: "macros.js"
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: [center, middle]
---

## Today's goals

- Introduction to Regular Expressions

- Understand the fundamentals of Web Scrapping

- Learn how to use an API

---

## Regular Expressions: What is it?

---

## Regular Expressions 101: Type of fields

Most common:

- `.` Any character
- `[0-9]` Numbers (without decimals or points)
- `[a-z]` Lower-case letters
- `[A-Z]` Upper-case letters
- `[a-zA-Z]` Lower or upper case letters.
- `[a-zA-Z0-9]` Any alpha-numeric

Other important characters

- `\s` space
- `^` beginning of the text
- `$` end of the text
- `|` or (logical or)

---

## Regular Expressions 101: Type of fields (cont. 1)

These usually come together with specifying how many times (repetition):

- `[...]?` Zero or one match.
- `[...]*` Zero or more matches
- `[...]+` One or more matches
- `[...]{n,}` At least `n` matches
- `[...]{,m}` at most `m` matches
- `[...]{n,m}` Between `n` and `m` matches.

Take a look at https://regex101.com/

---

## Regular Expressions 101: Type of fields (cont. 2)

A few examples:

```{r}
c("History 123")
```


---

## Regular Expressions 101: Common examples

1. Lookup text: `base::grepl()`, `stringr::str_match()`.

2. Replace text: `base::gsub()`, `stringr::str_replace()`.

3. Extract text: `base::regmatches()`, `stringr::str_extract()`.

---

## Data

This week we will continue using Textmining dataset

```{r load-data}
library(data.table)
library(stringr)

fn <- "mtsamples.csv"
if (!file.exists(fn))
  download.file(
    url = "https://github.com/USCbiostats/data-science-data/raw/master/00_mtsamples/mtsamples.csv",
    destfile = fn
  )
mtsamples <- fread(fn, sep = ",", header = TRUE)
```


---

## Regex Lookup Text: Tumor

We would like to see if this is tumor related entry. For that we can simply use
the following code:

```{r lookup-tumor}
# How many entries contain the word tumor
mtsamples[grepl("tumor", description, ignore.case = TRUE), .N] 

# Generating a column tagging tumor
mtsamples[, tumor_related := grepl("tumor", description)]

# Taking a look at a few examples
mtsamples[tumor_related == TRUE, .(description)][1:3,]
```

---

## Regex Lookup text: Gender of the patient

Now, let's try to guess the gender of the patient. To do so, we could tag by
using the words *he, she, him, her, his, hers* (see [this article on sexist text](https://dictionary.cambridge.org/grammar/british-grammar/sexist-language?q=He%2C+she%2C+him%2C+her%2C+his%2C+hers)):

```{r}
mtsamples[, gender := str_extract(
  string  = tolower(transcription),
  pattern = "he|his|him|she|hers|her"
)]
```

What is the problem with this approach?

---

## Regex Lookup text: Gender of the patient (cont. 1)

For this we use the following regular expression:

`(?<=\\W|^)(he|his|him|she|hers|her)(?=\\W|$)`

- `(?<=\\W|^)` Look back `(?<=)`, and in this case, look back for the beginning of the sentence `^`, or `|` a word `\W`.
- `he|his|him...` any of these words
- `(?=\\w|$)` Look forward `(?=)` 

```{r gender-regex}
# Lower case everything
mtsamples[, gender := tolower(transcription)]
mtsamples[, gender := str_extract(
  string  = gender, 
  pattern = "(?<=\\W|^)(he|his|him|she|hers|her)(?=\\W|$)"
  )]
mtsamples[1:10, gender]
```

---

## Regex Lookup text: Gender of the patient (cont. 2)

```{r gender-regex-cont}
mtsamples[, female := fifelse(gender %in% c("he", "his", "him"), FALSE, TRUE)]
mtsamples[, table(gender, female, useNA = "always")]
```


---

## Regex Extract Text: Type of Cancer

- Imagine now that you need to see the types of Cancer mentioned in the data.

- For simplicity, let's assume that, if specified, it is in the form of `TYPE cancer`, i.e. single word.

- We are interested in the word before cancer, how can we capture this?

---

## Regex Extract Text: Type of Cancer

We can just try to **extract** the phrase `"[some word] cancer"`, in particular, we could use the
following regular expression

`[a-zA-Z0-9-_]{4,}\\s*cancer`

Where

- `[a-zA-Z0-9-_]{4,}` captures any character in the ranges `a-zA-Z0-9` including `-` and `_`. 
   Furthermore, for this match to work there must be at least 4 characters,
- `\\s*` captures 0 or more white-spaces, and
- `cancer` captures the word cancer:

```{r cancer-regex}
mtsamples[, cancer_type := str_extract(tolower(keywords), "[a-zA-Z0-9-_]{4,}\\s*cancer")]
mtsamples[, table(cancer_type)]
```


---

## Fundamentals of Web Scrapping: What

Define what is is


---

## Fundamentals of Web Scrapping: How

- Direct data download v1 (download a plain-text-file)

- Raw HTML data (unstructured)

- Use an Application Programming Interface (API) (structured data)

---

## Web scraping with `xml2` and the `rvest` package

- The [`rvest`](https://cran.r-project.org/package=rvest) R package provides various tools for reading and processing web data.

- Under-the-hood, `rvest` is a wrapper of the [`xml2`](https://cran.r-project.org/package=xml2)
and [`httr`](https://cran.r-project.org/package=httr) R packages.


```{r setup-scrape}
library(rvest)

# Reading the HTML table
covid <- read_html(
  x = "https://en.wikipedia.org/wiki/COVID-19_pandemic_death_rates_by_country"
  )

# Let's the the output
covid
```

---

## Web scraping with `xml2` and the `rvest` package (cont. 1)

- We want to get the HTML table that shows up in the doc To do such, we can use the
  function `xml2::xml_find_all()` and `rvest::html_table()`

- The first will locate the place in the document that matches a given XPath
  expression.
  
- [XPath](), XML Path Language, is a query language to select nodes in a XML
  document.
  
- A nice tutorial can be found [here](https://www.w3schools.com/xml/xpath_intro.asp)

- Modern Web browsers make it easy to use XPath!

Live Example
  
---

## Web scraping with `xml2` and the `rvest` package (cont. 2)

Now that we know what is the path, let's use that and extract

```{r}
table <- xml_find_all(covid, xpath = "/html/body/div[3]/div[3]/div[5]/div[1]/table[2]")
table <- html_table(table) # This returns a list of tables
head(table[[1]])
```

```{r}
pb <- read_html("https://pubmed.ncbi.nlm.nih.gov/?term=flu&page=2")
pb <- xml_find_all(pb, xpath = "/html/body/main/div[9]/div[2]/div[2]/div[1]/span")
xml_text(pb)
```


---

## Setup for the APIs

For this part, we will be using the `httr()` package, which is a wrapper of the
`curl()` package, which in turn provides access to the `curl` library that
is used to communicate with APIs.

```{r setup-api}
library(httr)
```


---

Electronic Prescribing Adoption and Use by County

https://dashboard.healthit.gov/datadashboard/documentation/electronic-prescribing-adoption-use-data-documentation-county.php

```{r eval = FALSE}
ans <- GET(
  url    = "https://dashboard.healthit.gov/api/open-api.php",
  query  = list(source="AHA_2008-2015.csv", region="California", period=2020),
  config = config(connecttimeout = 1300)
)
```

---

## API Example 1: Using a Token

- Sometimes, APIs are not completely open, you need to register.

- In this example, I'm using a token which I obtained [here](https://www.ncdc.noaa.gov/cdo-web/token)

- You can find information about the [National Centers for Environmental Information](https://www.ncdc.noaa.gov/)
  API [here](https://www.ncdc.noaa.gov/cdo-web/webservices/v2)

---

## API Example 1: Using a Token (cont. 1)

- The way to pass the token will depend on the API service.

- Some require authentication, others need you to pass it as an argument of the query,
  i.e., directly in the URL.
  
- In this case, we pass it on the header.


```{r donwload-climate-data, eval = Sys.info()["login"] == "george", cache = TRUE}
token        <- readLines("vegayon-noaa-api-token.txt")
stations_api <- GET(
  url    = "https://www.ncdc.noaa.gov/cdo-web/api/v2/stations",
  config = add_headers(token = token),
  query  = list(limit = 1000)
)
```


This is equivalent to using the following query

```bash
curl --header "token: [YOURTOKEN]" \
  https://www.ncdc.noaa.gov/cdo-web/api/v2/stations?limit=1000
```

---

Again, we can recover the data using the `content()` function:

```{r download-climate-data-cont, eval = Sys.info()["login"] == "george"}
ans <- content(stations_api)
ans$results[[1]]
```

---

## API Example 2: Using a Token (cont. 2)



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
