---
title: "Assignment 4 - HPC and SQL"
link-citations: true
toc: false
---

## Due Date

This assignment is due by 11:59pm Pacific Time on Friday, November 21st, 2025.

The learning objectives are to write faster code for a computational task
requiring a loop and to implement some queries and basic data wrangling in SQL.

# HPC

For Questions 1 and 2, rewrite the provided R functions to make them faster. It
is OK (and recommended) to take a look at StackOverflow and Google.

Using the synthetic dataset provided, show that your new function returns the
same output as the original function. Then use `microbenchmark` to show that
your version is faster and provide a visualization of the runtimes.

### Question 1 (20 points)

 - `fun1`: Find the total for each row of a numeric matrix. Output should be a vector with length equal to the number of rows of the input matrix.

```{r, eval = FALSE}
# Total for each row
fun1 <- function(mat) {
  n <- nrow(mat)
  ans <- double(n) 
  for (i in 1:n) {
    ans[i] <- sum(mat[i, ])
  }
  ans
}

fun1alt <- function(mat) {
  # YOUR CODE HERE
}
```

Generate synthetic data (`dat`):

```{r, eval = FALSE}
# Use the functions with this data
set.seed(2315)
dat <- matrix(rnorm(200 * 100), nrow = 200)
```


### Question 2 (20 points)

 - `fun2`: Calculate the running (cumulative) total along each row of a numeric matrix. Output should be a matrix with dimensions equal to the input matrix.

```{r, eval = FALSE}
# running cumulative sum by row
fun2 <- function(mat) {
  n <- nrow(mat)
  k <- ncol(mat)
  ans <- mat
  for (i in 1:n) {
    for (j in 2:k) {
      ans[i,j] <- mat[i, j] + ans[i, j - 1]
    }
  }
  ans
}

fun2alt <- function(mat) {
  # YOUR CODE HERE
}
```

Use the same synthetic datset as above.

### Question 3 (20 points)

For this question, we will ask you to parallelize a process for computationally
estimating the value of $\pi$.

The following function allows us to estimate $\pi$ through a simple simulation:

```{r, eval = FALSE}
sim_pi <- function(n = 1000, i = NULL) {
  p <- matrix(runif(n*2), ncol = 2)
  mean(rowSums(p^2) < 1) * 4
}

# Here is an example of the run
set.seed(156)
sim_pi(1000) # 3.132
```

In order to get more accurate estimates, we can run this function multiple
times and take the mean:

```{r, eval = FALSE}
# This runs the simulation a 4,000 times, each with 10,000 points
set.seed(1231)
system.time({
  ans <- unlist(lapply(1:4000, sim_pi, n = 10000))
  print(mean(ans))
})
```

Rewrite the previous code using `parLapply()` (or your parallelization method of
choice) to parallelize it. Run the code once, using `system.time()`, to show
that your version is faster.

```{r, eval = FALSE}
# YOUR CODE HERE
system.time({
  # YOUR CODE HERE
  ans <- # YOUR CODE HERE
  print(mean(ans))
  # YOUR CODE HERE
})
```

Visualize your results in two ways: 

1. Show that both versions produce similar ranges of estimates for $\pi$.
2. Show that the parallel version runs faster.

# SQL

Setup a temporary database of movie data by running the following code chunk:

```{r, eval = FALSE}
# install.packages(c("RSQLite", "DBI"))

library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
film <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/film.csv")
film_category <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/film_category.csv")
category <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/category.csv")

# Copy data.frames to database
dbWriteTable(con, "film", film)
dbWriteTable(con, "film_category", film_category)
dbWriteTable(con, "category", category)
```

When you write a new chunk, remember to replace the `r` with `sql,
connection=con`. Some of these questions will require you to use an inner join.
Read more about them here [https://www.w3schools.com/sql/sql_join_inner.asp](https://www.w3schools.com/sql/sql_join_inner.asp)

Answer each of the questions below with SQL queries.

## Question 4 (10 points)

How many movies are available in each `rating` category?

## Question 5 (10 points)

What is the average replacement cost and rental rate for each `rating` category?

## Question 6 (10 points)

Use table `film_category` together with `film` to find how many films there are
with each category ID.

## Question 7 (10 points)

Incorporate the `category` table into the answer to the previous question to
find the name of the most popular category.
