---
title: "Lab 04 - Data Visualization"
output: html_document
date: 2021-09-16
link-citations: yes
---

```{r}
# Set to TRUE if you want the blocks to be evaluated 
EVAL <- FALSE
ECHO <- EVAL
```


```{r setup, message=FALSE, echo = ECHO, warning=FALSE, eval = EVAL}
#install.packages(c("data.table","leaflet"))
library(data.table)
library(leaflet)
library(tidyverse)
```

# Learning Goals

- Read in and prepare the meteorological dataset
- Create several graphs with different `geoms()` in `ggplot2`
- Create a facet graph
- Conduct some customizations to the graphs
- Create a more detailed map using `leaflet()`


# Lab Description

We will again work with the meteorological data presented in lecture.

**The objective of the lab is to examine the association between weekly average dew point temperature and wind speed in four regions of the US and by elevation.**

# Steps

### 1. Read in the data

First download and then read in with data.table:fread()

```{r, echo = TRUE, message=FALSE, eval = EVAL}
if (!file.exists("met_all.gz"))
  download.file(
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
    destfile = "met_all.gz",
    method   = "libcurl",
    timeout  = 60
    )
met <- data.table::fread("met_all.gz")
```

### 2. Prepare the data

- Remove temperatures less than -17C
- Make sure there are no missing data in the key variables coded as 9999, 999, etc
- Generate a date variable using the functions `as.Date()` (hint: You will need the following to create a date `paste(year, month, day, sep = "-")`).
- Using the `data.table::week` function, keep the observations of the first week of the month.
- Compute the mean by station of the variables `temp`, `rh`, `wind.sp`, `vis.dist`, `dew.point`, `lat`,
`lon`, and `elev`.
- Create a region variable for NW, SW, NE, SE based on lon = -98.00 and lat = 39.71 degrees
- Create a categorical variable for elevation as in the lecture slides

```{r, eval = EVAL, echo = ECHO}
met <- met[temp > -17][elev == 9999.0, elev := NA]
met[, week := week(as.Date(paste(year, month, day, sep = "-")))]
met <- met[week == min(week, na.rm = TRUE)]

met_avg <- met[,.(temp=mean(temp,na.rm=TRUE), rh=mean(rh,na.rm=TRUE), wind.sp=mean(wind.sp,na.rm=TRUE), 
                vis.dist=mean(vis.dist,na.rm=TRUE), dew.point = mean(dew.point, na.rm=TRUE), lat=mean(lat), lon=mean(lon), 
                elev=mean(elev,na.rm=TRUE)), by="USAFID"]

met_avg$elev_cat <- ifelse(met_avg$elev> 252, "high", "low")

met_avg$region <- ifelse(met_avg$lon > -98 & met_avg$lat >39.71, "north east",
                         ifelse(met_avg$lon > -98 & met_avg$lat < 39.71, "south east",
                                ifelse(met_avg$lon < -98 & met_avg$lat >39.71, "north west", "south west")))

table(met_avg$region)

```

### 3. Use `geom_violin` to examine the wind speed and dew point temperature by region

You saw how to use `geom_boxplot` in class. Try using `geom_violin` instead (take a look at the help).
(hint: You will need to set the `x` aesthetic to 1)

- Use facets
- Make sure to deal with `NA` category
- Describe what you observe in the graph

```{r, eval = EVAL, echo = ECHO}
met_avg %>%
  filter(!(region %in% NA)) %>%
ggplot()+
  geom_violin(mapping = aes(y=wind.sp, x=1)) +
  facet_wrap(~region, nrow=2)

met_avg %>%
  filter(!(region %in% NA)) %>%
ggplot()+
  geom_boxplot(mapping = aes(y=rh, fill=region)) +
  facet_wrap(~region, nrow=2)
```


### 4. Use `geom_jitter` with `stat_smooth` to examine the association between dew point temperature and wind speed by region

- Colour points by region
- Make sure to deal with `NA` category
- Fit a linear regression line by region
- Describe what you observe in the graph

```{r, eval = EVAL, echo = ECHO}
met_avg %>%
filter(!(region %in% NA)) %>%
  ggplot(mapping = aes(x=dew.point, y=rh, color=region))+
  geom_jitter() + 
  stat_smooth(method=lm)
  
```


### 5. Use `geom_bar` to create barplots of the weather stations by elevation category coloured by region

- Bars by elevation category using `position="dodge"`
- Change colours from the default. Colour by region using `scale_fill_brewer` see [this](http://rstudio-pubs-static.s3.amazonaws.com/5312_98fc1aba2d5740dd849a5ab797cc2c8d.html)
- Create nice labels on axes and add a title
- Describe what you observe in the graph
- Make sure to deal with `NA`

```{r warning=FALSE, message=FALSE, eval = EVAL, echo = ECHO}
met_avg %>%
filter(!(region %in% NA)) %>%
  ggplot()+
  geom_bar(mapping=aes(x=elev_cat,fill=region), position = "dodge")+
  scale_fill_brewer(palette = "PuOr")+
  labs(title="Number of weather stations by elevation category and region", x="Elevation Category", y= "Count")+
  theme_bw()

```

### 6. Use `stat_summary` to examine mean dew point and wind speed by region with standard deviation error bars

- Make sure to remove `NA`
- Use fun.data="mean_sdl" in `stat_summary`
- Add another layer of `stats_summary` but change the geom to `"errorbar"` (see the help).
- Describe the graph and what you observe

```{r, eval = EVAL, echo = ECHO}
met_avg %>%
filter(!(region %in% NA)) %>%
  ggplot(mapping=aes(x=region, y=dew.point)) +
  stat_summary(fun.data="mean_sdl", geom="errorbar") +
  stat_summary(fun.data="mean_sdl")
```

- Dew point temperature is...
- Wind speed is...

### 7. Make a map showing the spatial trend in relative h in the US

- Make sure to remove `NA`
- Use leaflet()
- Make a colour palette with custom colours
- Use `addMarkers` to include the top 10 places in relative h (hint: this will be useful `rank(-rh) <= 10`)
- Add a legend

```{r, eval = EVAL, echo = ECHO}
met_avg2<-met_avg[!is.na(rh)]

# Top five
top5 <- met_avg2[rank(-rh) <= 10]

rh_pal = colorNumeric(c('blue','purple','red'), domain=met_avg2$rh)
leaflet(met_avg2) %>%
  addProviderTiles('OpenStreetMap') %>%
  addCircles(lat=~lat, lng=~lon, color=~rh_pal(rh), label=~paste0(round(rh,2), ' rh'), opacity=1,fillOpacity=1, radius=500) %>%
  addMarkers(lat=~lat, lng=~lon, label=~paste0(round(rh,2), ' rh'), data = top5) %>%
  addLegend('bottomleft',pal=rh_pal, values=met_avg2$rh, title="Relative Humidity", opacity=1)

```

- Describe trend in RH across the US

### 8. Use a ggplot extension

- Pick and extension (except cowplot) from [here](https://exts.ggplot2.tidyverse.org/gallery/) and make a plot of your choice using the met data (or met_avg)
- Might want to try examples that come with the extension first (e.g. ggtech, gganimate, ggforce)

