---
title: "Lab 04 - Exploratory Data Analysis"
output: html_document
link-citations: yes
---



```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
#install.packages(c("data.table","leaflet"))
library(data.table)
library(leaflet)
library(tidyverse)
```

# Learning Goals

- Read in and prepare the meteorological dataset
- Create several graphs with different `geoms()` in `ggplot2`
- Create a facet graph
- Conduct some customizations to the graphs
- Create a more detailed map using `leaflet()`


# Lab Description

We will again work with the meteorological data presented in lecture.

**The objective of the lab is to examine the association between monthly average dew point temperature and relative humidity in four regions of the US and by elevation.**

# Steps

### 1. Read in the data

First download and then read in with data.table:fread()

```{r, echo=TRUE, message=FALSE}
download.file("https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz", "met_all.gz", method="libcurl", timeout = 60)
met <- data.table::fread("met_all.gz")
```

### 2. Prepare the data

- Remove temperatures less than -17C
- Make sure there are no missing data in the key variables coded as 9999, 999, etc
- Take monthly averages by weather station
- Create a region variable for NW, SW, NE, SE based on lon = -98.00 and lat = 39.71 degrees
- Create a categorical variable for elevation as in the lecture slides

```{r}

```

### 3. Use `geom_boxplot` to examine the dew point temperature and relative humidity by region

- Use facets
- Make sure to deal with `NA` category
- Describe what you observe in the graph
```{r}

```

### 4. Use `geom_point` with `stat_smooth` to examine the association between dew point temperature and relative humidity by region

- Colour points by region
- Make sure to deal with `NA` category
- Fit a linear regression line by region
- Describe what you observe in the graph

```{r}

```

### 5. Use `geom_bar` to create barplots of the weather stations by elevation category coloured by region

- Bars by elevation category
- Change colours from the default. Colour by region using `scale_fill_brewer` see [this](http://rstudio-pubs-static.s3.amazonaws.com/5312_98fc1aba2d5740dd849a5ab797cc2c8d.html)
- Create nice labels on axes and add a title
- Describe what you observe in the graph

```{r warning=FALSE, message=FALSE}

```

### 6. Use `stat_summary` to examine mean dew point and relative humidity by region with standard deviation error bars

- Make sure to remove `NA`
- Use fun.data="mean_sdl" in `stat_summary`
- Describe the graph and what you observe

```{r}

```

### 7. Make a map showing the spatial trend in relative humidity in the US

- Make sure to remove `NA`
- Use leaflet()
- Make a colour palette with custom colours
- Add a legend

```{r}

```

### 8. Use a ggplot extension

- Pick and extension (except cowplot) from [here](https://exts.ggplot2.tidyverse.org/gallery/) and make a plot of your choice using the met data (or met_avg)
- Might want to try examples that come with the extension first (e.g. ggtech, gganimate, ggforce)

