---
title: "Lab 07 - Web scraping and Regular Expressions"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(include  = TRUE)
```


# Learning goals

- Use a real world API to make queries and process the data.
- Use regular expressions to parse the information.
- Practice your GitHub skills.

# Lab description

In this lab, we will be working with the [NCBI API](https://www.ncbi.nlm.nih.gov/home/develop/api/)
to make queries and extract information using XML and regular expressions. For this lab, we will
be using the `httr`, `xml2`, and `stringr` R packages.

This markdown document should be rendered using `github_document` document.

## Question 1: How many sars-cov-2 papers?

Build an automatic counter of sars-cov-2 papers using PubMed. You will need to apply XPath as we did during the lecture to extract the number of results returned by PubMed in the following web address:

```
https://pubmed.ncbi.nlm.nih.gov/?term=sars-cov-2
```

Complete the lines of code:

```{r counter-pubmed, eval=FALSE}
# Downloading the website
website <- xml2::read_html("[URL]")

# Finding the counts
counts <- xml2::xml_find_first(website, "[XPath]")

# Turning it into text
counts <- as.character(counts)

# Extracting the data using regex
stringr::str_extract(counts, "[REGEX FOR NUMBERS WITH COMMAS/DOTS]")
```

Don't forget to commit your work!

## Question 2: Academic publications on COVID19 and Hawaii

You need to query the following
The parameters passed to the query are documented [here](https://www.ncbi.nlm.nih.gov/books/NBK25499/).

Use the function `httr::GET()` to make the following query:

1. Baseline URL: https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi

2. Query parameters:

    - db: pubmed
    - term: covid19 hawaii
    - retmax: 1000


```{r papers-covid-hawaii, eval=FALSE}
library(httr)
query_ids <- GET(
  url   = "BASELINE URL",
  query = list("QUERY PARAMETERS")
)

# Extracting the content of the response of GET
ids <- httr::content(query_ids)
```

The query will return an XML object, we can turn it into a character list to
analyze the text directly with `as.character()`. Another way of processing the
data could be using lists with the function `xml2::as_list()`. We will skip the
latter for now.

Take a look at the data, and continue with the next question (don't forget to 
commit and push your results to your GitHub repo!).

## Question 3: Get details about the articles

The Ids are wrapped around text in the following way: `<Id>... id number ...</Id>`.
we can use a regular expression that extract that information. Fill out the
following lines of code:

```{r get-ids, eval = FALSE}
# Turn the result into a character vector
ids <- as.character(ids)

# Find all the ids 
ids <- stringr::str_extract_all(ids, "PATTERN")[[1]]

# Remove all the leading and trailing <Id> </Id>. Make use of "|"
ids <- stringr::str_remove_all(ids, "PATTERN")
```

With the ids in hand, we can now try to get the abstracts of the papers. As
before, we will need to coerce the contents (results) to a list using:

1. Baseline url: https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi

2. Query parameters:
    - db: pubmed
    - id: A character with all the ids separated by comma, e.g., "1232131,546464,13131"
    - retmax: 1000
    - rettype: abstract
    
**Pro-tip**: If you want `GET()` to take some element literal, wrap it around `I()` (as you would do in a formula in R). For example, the text `"123,456"` is replaced with `"123%2C456"`. If you don't want that behavior, you would need to do the following `I("123,456")`.
    
```{r get-abstracts, eval = FALSE}
publications <- GET(
  url   = "BASELINE URL HERE",
  query = list(
    "PARAMETERS OF THE QUERY"
    )
)

# Turning the output into character vector
publications <- httr::content(publications)
publications_txt <- as.character(publications)
```

With this in hand, we can now analyze the data. This is also a good time for committing and pushing your work!

## Question 4: Distribution of universities, schools, and departments

Using the function `stringr::str_extract_all()` applied on `publications_txt`, capture all the terms of the form:

1.    University of ...
2.    ... Institute of ...

Write a regular expression that captures all such instances

```{r univ-institute-regex, eval = FALSE}
institution <- str_extract_all(
  publications_txt,
  "[YOUR REGULAR EXPRESSION HERE]"
  ) 
institution <- unlist(institution)
table(institution)
```

Repeat the exercise and this time focus on schools and departments in the form of

1.    School of ...
2.    Department of ...

And tabulate the results

```{r school-department, eval = FALSE}
schools_and_deps <- str_extract_all(
  abstracts_txt,
  "[YOUR REGULAR EXPRESSION HERE]"
  )
table(schools_and_deps)
```

## Question 5: Form a database

We want to build a dataset which includes the title and the abstract of the
paper. The title of all records is enclosed by the HTML tag `ArticleTitle`, and
the abstract by `Abstract`. 

Before applying the functions to extract text directly, it will help to process
the XML a bit. We will use the `xml2::xml_children()` function to keep one element
per id. This way, if a paper is missing the abstract, or something else, we will be able to properly match PUBMED IDS with their corresponding records.


```{r one-string-per-response, eval = FALSE}
pub_char_list <- xml2::xml_children(publications)
pub_char_list <- sapply(pub_char_list, as.character)
```

Now, extract the abstract and article title for each one of the elements of
`pub_char_list`. You can either use `sapply()` as we just did, or simply
take advantage of vectorization of `stringr::str_extract`

```{r extracting-last-bit, eval = FALSE}
abstracts <- str_extract(pub_char_list, "[YOUR REGULAR EXPRESSION]")
abstracts <- str_remove_all(abstracts, "[CLEAN ALL THE HTML TAGS]")
abstracts <- str_remove_all(abstracts, "[CLEAN ALL EXTRA WHITE SPACE AND NEW LINES]")
```

How many of these don't have an abstract? Now, the title

```{r process-titles, eval = FALSE}
titles <- str_extract(pub_char_list, "[YOUR REGULAR EXPRESSION]")
titles <- str_remove_all(titles, "[CLEAN ALL THE HTML TAGS]")
```

Finally, put everything together into a single `data.frame` and use
`knitr::kable` to print the results

```{r build-db, eval = FALSE}
database <- data.frame(
  "[DATA TO CONCATENATE]"
)
knitr::kable(database)
```

Done! Knit the document, commit, and push.

## Final Pro Tip (optional)

You can still share the HTML document on github. You can include a link in your `README.md` file as the following:

```md
View [here](https://ghcdn.rawgit.org/:user/:repo/:tag/:file)
```

For example, if we wanted to add a direct link the HTML page of lecture 7, we could do something like the following:

```md
View [here](https://ghcdn.rawgit.org/USCbiostats/PM566/master/static/slides/07-apis-regex/slides.html)
```


