---
title: "Lab 07 - Web scraping and Regular Expressions"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(eval = FALSE, include  = FALSE)
```


# Learning goals

- Use the `merge()` function to join two datasets.
- Deal with missings and impute data.
- Identify relevant observations using `quantile()`.
- Practice your GitHub skills.

# Lab description

For this lab we will be, again, dealing with the meteorological dataset downloaded from the NOAA, the `met`. In this case, we will use `data.table` to answer some questions regarding the `met` dataset, while at the same time practice your Git+GitHub skills for this project.

This markdown document should be rendered using `github_document` document.

# Part 1: Setup the Git project and the GitHub repository

1. Go to your documents (or wherever you are planning to store the data) in your computer, and create a folder for this project, for example, "PM566-labs"

2. In that folder, save [this template](https://raw.githubusercontent.com/USCbiostats/PM566/master/content/assignment/05-lab.Rmd) as "README.Rmd". This will be the markdown file where all the magic will happen.

3. Go to your GitHub account and create a new repository, hopefully of the same name that this folder has, i.e., "PM566-labs".

4. Initialize the Git project, add the "README.Rmd" file, and make your first commit.

5. Add the repo you just created on GitHub.com to the list of remotes, and push your commit to origin while setting the upstream.

Most of the steps can be done using command line:

```sh
# Step 1
cd ~/Documents
mkdir PM566-labs
cd PM566-labs

# Step 2
wget https://raw.githubusercontent.com/USCbiostats/PM566/master/content/assignment/05-lab.Rmd 
mv 05-lab.Rmd README.md

# Step 3
# Happens on github

# Step 4
git init
git add README.Rmd
git commit -m "First commit"

# Step 5
git remote add origin git@github.com:[username]/PM566-labs
git push -u origin master
```

You can also complete the steps in R (replace with your paths/username when needed)

```r
# Step 1
setwd("~/Documents")
dir.create("PM566-labs")
setwd("PM566-labs")

# Step 2
download.file(
  "https://raw.githubusercontent.com/USCbiostats/PM566/master/content/assignment/05-lab.Rmd",
  destfile = "README.Rmd"
  )

# Step 3: Happens on Github

# Step 4
system("git init && git add README.Rmd")
system('git commit -m "First commit"')

# Step 5
system("git remote add origin git@github.com:[username]/PM566-labs")
system("git push -u origin master")
```

Once you are done setting up the project, you can now start working with the MET data.

## Setup in R

1. Load the `data.table` (and the `dtplyr` and `dplyr` packages if you plan to work with those).

2. Load the met data from https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz, and also the station data. For the later, you can use the code we used during lecture to pre-process the stations data:

```{r stations-data, eval = FALSE}
# Download the data
stations <- fread("ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
stations[, USAF := as.integer(USAF)]

# Dealing with NAs and 999999
stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

# Selecting the three relevant columns, and keeping unique records
stations <- unique(stations[, list(USAF, CTRY, STATE)])

# Dropping NAs
stations <- stations[!is.na(USAF)]

# Removing duplicates
stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]
```

3. Merge the data as we did during the lecture.

## Question 1: Academic publications on COVID19 and Hawaii

You need to query the following
The parameters passed to the query are documented [here](https://www.ncbi.nlm.nih.gov/books/NBK25499/).

Use the function `httr::GET()` to make the following query:

1. Baseline URL: https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi

2. Query parameters:

    - db: pubmed
    - term: covid19 hawaii


```{r}
library(httr)
query_ids <- GET(
  url   = "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi",
  query = list(
    db     = "pubmed",
    term   = "covid19 hawaii",
    retmax = 1000
  )
)
query_ids
```

The query will return an XML object, we can turn it into a list using the function
`xml2::as_list()`. And from there, access the relevant output, which is the list
of IDs:

```{r get-ids}
# Processing the ids
ids <- xml2::as_list(content(query_ids))
ids <- unique(unlist(ids$eSearchResult$IdList, recursive = TRUE))
```

With the ids in hand, we can now try to get the abstracts of the papers. As
before, we will need to coerce the contents (results) to a list using `as_list()`


```{r get-abstracts}
query_details <- GET(
  url   = "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi",
  query = list(
    db     = "pubmed",
    id     = I(paste(ids, collapse = ",")),
    retmax = 10000,
    rettype = "abstract"
  )
)

abstracts <- content(query_details)
abstracts <- xml2::as_list(abstracts)
```

```{r abstract-processing}
# How many abstracts are out there?
length(abstracts$PubmedArticleSet)

# Extracting the abstracts only
abstracts_txt <- character(length(abstracts$PubmedArticleSet))
authors       <- abstracts_txt
for (i in 1:length(abstracts_txt)) {
  
  # Extracting the abstract ---------------------------------------------------
  tmp <- abstracts$PubmedArticleSet[[i]]$MedlineCitation$Article$Abstract$AbstractText[[1]]
  
  # Is it length greater than zero?
  if (length(tmp))
    abstracts_txt[i] <- unlist(tmp, recursive = TRUE)
  
  # Extracting the authors ----------------------------------------------------
  tmp <- abstracts$PubmedArticleSet[[i]]$MedlineCitation$Article$AuthorList
  tmp <- lapply(tmp, "[[", "AffiliationInfo")
  authors[i] <- paste(unlist(tmp, recursive = TRUE), collapse = ";")
}

```

## Question 2: Distribution of universities, schools, and departments

Using the function `stringr::str_extract_all()` applied on `authors`, capture all the terms of the form:

1.    University of ...
2.    ... Institute of ...

```{r}
institution <- stringr::str_extract_all(
  authors,
  "University of [a-zA-Z]+|[a-zA-Z]+ Institute of [a-zA-Z]+"
  ) 
institution <- unlist(institution)
table(institution)
```

Repeat the exercise and this time focus on schools and departments

1.    School of ...
2.    Department of ...

```{r}
stringr::str_extract_all(
  authors,
  "Department of [a-zA-Z]+|University of [a-zA-Z]+|Division of [a-zA-Z]+|School of [a-zA-Z]+|[a-zA-Z]+ Institute of [a-zA-Z]+"
  )
```

## Question 3: 


