---
title: "Lab 11 - Interactive Visualization"
output: html_document
link-citations: yes
editor_options: 
  chunk_output_type: console
---



```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(knitr)

opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  #eval=FALSE,
  echo = TRUE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px")

```

# Learning Goals

- Read in and process the COVID dataset from the New York Times GitHub repository
- Create interactive graphs of different types using `plot_ly()` and `ggplotly()` functions
- Customize the hoverinfo and other plot features
- Create a Choropleth map using `plot_geo()`
- Create an interactive table using `DataTable`

# Lab Description

We will work with the COVID data presented in lecture. Recall the dataset consists of COVID-19 cases and deaths in each US state during the course of the COVID epidemic.

**The objective of this lab is to identify how long after cases increased deaths increased, and after cases decreased deaths decreased, and plot data to demonstrate this**

# Steps

## I. Reading and processing the New York Times (NYT) state-level COVID-19 data

### 1. Read in the data

- Read in the COVID data with data.table:fread() from the NYT GitHub repository: "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv"
- Read in the state population data with data.table:fread() from the repository: "https://raw.githubusercontent.com/COVID19Tracking/associated-data/master/us_census_data/us_census_2018_population_estimates_states.csv""
- Merge datasets

```{r, echo=TRUE, message=FALSE}

## data extracted from New York Times state-level data from NYT Github repository
# https://github.com/nytimes/covid-19-data

## state-level population information from us_census_data available on GitHub repository:
# https://github.com/COVID19Tracking/associated-data/tree/master/us_census_data

# load COVID state-level data from NYT
cv_states <- as.data.frame(data.table::fread("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv"))

# load state population data
state_pops <- as.data.frame(data.table::fread("https://raw.githubusercontent.com/COVID19Tracking/associated-data/master/us_census_data/us_census_2018_population_estimates_states.csv"))
state_pops$abb <- state_pops$state
state_pops$state <- state_pops$state_name
state_pops$state_name <- NULL
cv_states <- merge(cv_states, state_pops, by="state")
```

### 2. Look at the data

- Inspect the dimensions, `head`, and `tail` of the data
- Inspect the structure of each variables. Are they in the correct format?

```{r}
dim(cv_states)
head(cv_states)
tail(cv_states)
str(cv_states)
```

### 3. Format the data

- Make date into a date variable
- Make state into a factor variable
- Order the data first by state, second by date
- Confirm the variables are now correctly formatted
- Inspect the range values for each variable. What is the date range? The range of cases and deaths?

```{r}
# format the date
cv_states$date <- as.Date(cv_states$date, format="%Y-%m-%d")

# format the state and state abbreviation (abb) variables
state_list <- unique(cv_states$state)
cv_states$state <- factor(cv_states$state, levels = state_list)
abb_list <- unique(cv_states$abb)
cv_states$abb <- factor(cv_states$abb, levels = abb_list)

# order the data first by state, second by date
cv_states = cv_states[order(cv_states$state, cv_states$date),]

# Confirm the variables are now correctly formatted
str(cv_states)
head(cv_states)
tail(cv_states)

# Inspect the range values for each variable. What is the date range? The range of cases and deaths?
head(cv_states)
summary(cv_states)
min(cv_states$date)
max(cv_states$date)
```

### 4. Add `new_cases` and `new_deaths` and correct outliers

- Add variables for new cases, `new_cases`, and new deaths, `new_deaths`: 
  - Hint: `new_cases` is equal to the difference between cases on date i and date i-1, starting on date i=2

- Use `plotly` for EDA: See if there are outliers or values that don't make sense for `new_cases` and `new_deaths`. Which states and which dates have strange values?

- Correct outliers: Set negative values for `new_cases` or `new_deaths` to 0

- Recalculate `cases` and `deaths` as cumulative sum of updates `new_cases` and `new_deaths`

```{r}
# Add variables for new_cases and new_deaths:
for (i in 1:length(state_list)) {
  cv_subset = subset(cv_states, state == state_list[i])
  cv_subset = cv_subset[order(cv_subset$date),]

  # add starting level for new cases and deaths
  cv_subset$new_cases = cv_subset$cases[1]
  cv_subset$new_deaths = cv_subset$deaths[1]

  for (j in 2:nrow(cv_subset)) {
    cv_subset$new_cases[j] = cv_subset$cases[j] - cv_subset$cases[j-1]
    cv_subset$new_deaths[j] = cv_subset$deaths[j] - cv_subset$deaths[j-1]
  }

  # include in main dataset
  cv_states$new_cases[cv_states$state==state_list[i]] = cv_subset$new_cases
  cv_states$new_deaths[cv_states$state==state_list[i]] = cv_subset$new_deaths
}

# Inspect outliers using plotly
p1<-ggplot(cv_states, aes(x = date, y = new_cases, color = state)) + geom_line() + geom_point(size = .5, alpha = 0.5)
ggplotly(p1)
p1<-NULL # to clear from workspace

p2<-ggplot(cv_states, aes(x = date, y = new_deaths, color = state)) + geom_line() + geom_point(size = .5, alpha = 0.5)
ggplotly(p2)
p2<-NULL # to clear from workspace

# set negative new case or death counts to 0
cv_states$new_cases[cv_states$new_cases<0] = 0
cv_states$new_deaths[cv_states$new_deaths<0] = 0

# Recalculate `cases` and `deaths` as cumulative sum of updates `new_cases` and `new_deaths`
for (i in 1:length(state_list)) {
  cv_subset = subset(cv_states, state == state_list[i])

  # add starting level for new cases and deaths
  cv_subset$cases = cv_subset$cases[1]
  cv_subset$deaths = cv_subset$deaths[1]

  for (j in 2:nrow(cv_subset)) {
    cv_subset$cases[j] = cv_subset$new_cases[j] + cv_subset$cases[j-1]
    cv_subset$deaths[j] = cv_subset$new_deaths[j] + cv_subset$deaths[j-1]
  }
  # include in main dataset
  cv_states$cases[cv_states$state==state_list[i]] = cv_subset$cases
  cv_states$deaths[cv_states$state==state_list[i]] = cv_subset$deaths
}
```

### 5. Add additional variables

- Add population-normalized (by 100,000) variables for each variable type (rounded to 1 decimal place). Make sure the variables you calculate are in the correct format (`numeric`). You can use the following variable names:
  - `per100k` = cases per 100,000 population
  - `newper100k`= new cases per 100,000
  - `deathsper100k` = deaths per 100,000
  - `newdeathsper100k` = new deaths per 100,000

- Add a "naive CFR" variable representing `deaths / cases` on each date for each state

- Create a dataframe representing values on the most recent date, `cv_states_today`, as done in lecture

```{r}

# add population normalized (by 100,000) counts for each variable
cv_states$per100k =  as.numeric(format(round(cv_states$cases/(cv_states$population/100000),1),nsmall=1))
cv_states$newper100k =  as.numeric(format(round(cv_states$new_cases/(cv_states$population/100000),1),nsmall=1))
cv_states$deathsper100k =  as.numeric(format(round(cv_states$deaths/(cv_states$population/100000),1),nsmall=1))
cv_states$newdeathsper100k =  as.numeric(format(round(cv_states$new_deaths/(cv_states$population/100000),1),nsmall=1))

# add a naive_CFR variable = deaths / cases
cv_states = cv_states %>% mutate(naive_CFR = round((deaths*100/cases),2))

# create a `cv_states_today` variable
cv_states_today = subset(cv_states, date==max(cv_states$date))

```

## II. Scatterplots

### 6. Explore scatterplots using `plot_ly()`

- Create a scatterplot using `plot_ly()` representing `pop_density` vs. various variables (e.g. `cases`, `per100k`, `deaths`, `deathsper100k`) for each state on most recent date (`cv_states_today`)
  - Use hover to identify any outliers. 
  - Remove those outliers and replot.
- Choose one plot. For this plot:
  - Add hoverinfo specifying the state name, cases per 100k, and deaths per 100k, similarly to how we did this in the lecture notes
  - Add layout information to title the chart and the axes
  - Enable `hovermode = "compare"`

```{r}

# pop_density vs. cases
cv_states_today %>% 
  plot_ly(x = ~pop_density, y = ~cases, 
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70), marker = list(sizemode='diameter', opacity=0.5))

# filter out "District of Columbia"
cv_states_today_scatter <- cv_states_today %>% filter(state!="District of Columbia")

# pop_density vs. cases after filtering
cv_states_today_scatter %>% filter(state!="District of Columbia") %>% 
  plot_ly(x = ~pop_density, y = ~cases, 
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70), marker = list(sizemode='diameter', opacity=0.5))

# pop_density vs. deathsper100k
cv_states_today_scatter %>% filter(state!="District of Columbia") %>% 
  plot_ly(x = ~pop_density, y = ~deathsper100k, z = ~population,
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70), marker = list(sizemode='diameter', opacity=0.5))

# Adding hoverinfo
cv_states_today_scatter %>% 
  plot_ly(x = ~pop_density, y = ~deathsper100k,
          type = 'scatter', mode = 'markers', color = ~state,
          size = ~population, sizes = c(5, 70), marker = list(sizemode='diameter', opacity=0.5),
          hoverinfo = 'text',
          text = ~paste( paste(state, ":", sep=""), paste(" Cases per 100k: ", per100k, sep="") , paste(" Deaths per 100k: ",
                        deathsper100k, sep=""), sep = "<br>")) %>%
  layout(title = "Population-normalized COVID-19 deaths (per 100k) vs. population density for US states",
                  yaxis = list(title = "Deaths per 100k"), xaxis = list(title = "Population Density"),
         hovermode = "compare")
```

### 7. Explore scatterplot trend interactively using `ggplotly()` and `geom_smooth()`

- For `pop_density` vs. `newdeathsper100k` create a chart with the same variables using `gglot_ly()`
- Explore the pattern between $x$ and $y$ using `geom_smooth()`
  - Explain what you see. Do you think `pop_density` is a correlate of `newdeathsper100k`?

```{r}
p <- ggplot(cv_states_today_scatter, aes(x=pop_density, y=deathsper100k, size=population)) + geom_point() + geom_smooth()
ggplotly(p)
```


### 8. Multiple line chart

- Create a line chart of the `naive_CFR` for all states over time using `plot_ly()`
  - Use hoverinfo to identify states that had a "first peak"
  - Use the zoom and pan tools to inspect the `naive_CFR` for the states that had a "first peak" in September. How have they changed over time?
- Create one more line chart, for Texas only, which shows `new_cases` and `new_deaths` together in one plot. Hint: use `add_layer()`
  - Use hoverinfo to "eyeball" the approximate peak of deaths and peak of cases. What is the time delay between the peak of cases and the peak of deaths?

```{r}

# Line chart for naive_CFR for all states over time using `plot_ly()`
plot_ly(cv_states, x = ~date, y = ~naive_CFR, color = ~state, type = "scatter", mode = "lines")

# Line chart for Texas showing new_cases and new_deaths together
cv_states %>% filter(state=="Texas") %>% plot_ly(x = ~date, y = ~new_cases, type = "scatter", mode = "lines") %>% add_lines(x = ~date, y = ~new_deaths, type = "scatter", mode = "lines") 

```


### 9. Heatmaps

Create a heatmap to visualize `new_cases` for each state on each date greater than April 1st, 2020
- Start by mapping selected features in the dataframe into a matrix using the **tidyr** package function `pivot_wider()`, naming the rows and columns, as done in the lecture notes
- Use `plot_ly()` to create a heatmap out of this matrix
- Create a second heatmap in which the pattern of `new_cases` for each state over time becomes more clear by filtering to only look at dates every two weeks 

```{r}

# Map state, date, and new_cases to a matrix
library(tidyr)
cv_states_mat <- cv_states %>% select(state, date, new_cases) %>% filter(date>as.Date("2020-04-01"))
cv_states_mat2 <- as.data.frame(pivot_wider(cv_states_mat, names_from = state, values_from = new_cases))
rownames(cv_states_mat2) <- cv_states_mat2$date
cv_states_mat2$date <- NULL
cv_states_mat2 <- as.matrix(cv_states_mat2)

# Create a heatmap using plot_ly()
plot_ly(x=colnames(cv_states_mat2), y=rownames(cv_states_mat2),
             z=~cv_states_mat2,
             type="heatmap",
             showscale=T)

# Create a second heatmap after filtering to only include dates every other week
filter_dates <- seq(as.Date("2020-04-01"), as.Date("2020-10-01"), by="2 weeks")

cv_states_mat <- cv_states %>% select(state, date, new_cases) %>% filter(date %in% filter_dates)
cv_states_mat2 <- as.data.frame(pivot_wider(cv_states_mat, names_from = state, values_from = new_cases))
rownames(cv_states_mat2) <- cv_states_mat2$date
cv_states_mat2$date <- NULL
cv_states_mat2 <- as.matrix(cv_states_mat2)

# Create a heatmap using plot_ly()
plot_ly(x=colnames(cv_states_mat2), y=rownames(cv_states_mat2),
             z=~cv_states_mat2,
             type="heatmap",
             showscale=T)
```

### 10. Map

- Create a map to visualize the `naive_CFR` by state on May 1st, 2020
- Compare with a map visualizing the `naive_CFR` by state on most recent date
- Plot the two maps together using `subplot()`. Make sure the shading is for the same range of values (google is your friend for this)
- Describe the difference in the pattern of the CFR.

```{r}

### For May 1 2020

# Extract the data for each state by its abbreviation
cv_CFR <- cv_states %>% filter(date=="2020-05-01") %>% select(state, abb, naive_CFR, cases, deaths) # select data
cv_CFR$state_name <- cv_CFR$state
cv_CFR$state <- cv_CFR$abb
cv_CFR$abb <- NULL

# Create hover text
cv_CFR$hover <- with(cv_CFR, paste(state_name, '<br>', "CFR: ", naive_CFR, '<br>', "Cases: ", cases, '<br>', "Deaths: ", deaths))

# Set up mapping details
set_map_details <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)

# Make sure both maps are on the same color scale
shadeLimit <- 9

# Create the map
fig <- plot_geo(cv_CFR, locationmode = 'USA-states') %>% 
  add_trace(
    z = ~naive_CFR, text = ~hover, locations = ~state,
    color = ~naive_CFR, colors = 'Purples'
  )
fig <- fig %>% colorbar(title = "CFR May 1 2020", limits = c(0,shadeLimit))
fig <- fig %>% layout(
    title = paste('CFR by State as of', Sys.Date(), '<br>(Hover for value)'),
    geo = set_map_details
  )
fig_May1 <- fig

#############
### For Today

# Extract the data for each state by its abbreviation
cv_CFR <- cv_states_today %>%  select(state, abb, naive_CFR, cases, deaths) # select data
cv_CFR$state_name <- cv_CFR$state
cv_CFR$state <- cv_CFR$abb
cv_CFR$abb <- NULL

# Create hover text
cv_CFR$hover <- with(cv_CFR, paste(state_name, '<br>', "CFR: ", naive_CFR, '<br>', "Cases: ", cases, '<br>', "Deaths: ", deaths))

# Set up mapping details
set_map_details <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)

# Create the map
fig <- plot_geo(cv_CFR, locationmode = 'USA-states') %>% 
  add_trace(
    z = ~naive_CFR, text = ~hover, locations = ~state,
    color = ~naive_CFR, colors = 'Purples'
  )
fig <- fig %>% colorbar(title = "CFR May 1 2020", limits = c(0,shadeLimit))
fig <- fig %>% layout(
    title = paste('CFR by State as of', Sys.Date(), '<br>(Hover for value)'),
    geo = set_map_details
  )
fig_Today <- fig


### Plot together 
subplot(fig_May1, fig_Today)

```


